---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  warning = FALSE, message = FALSE
)
```


# PICmodel

<!-- badges: start -->
<!-- badges: end -->

This `R` package fits prevalence-incidence-clearance (PIC) models to interval-censored data to estimate the cumulative disease risk in a population with a temporarily elevated risk of disease, e.g., risk of CIN2+ in HPV-positive women. 

In longitudinal screening studies it is possible to observe prevalent, early or late events during follow-up. In our model, early events are modelled via a competing risks framework, infections either progress to the disease state or to a (latent) “cure” state (i.e. viral clearance) at constant rates. Late events are modelled by adding background risk to the model. Parameters can depend on individual risk factors and are estimated with an expectation-maximisation (EM) algorithm with weakly informative Cauchy priors. More details are given in the accompanying paper: *eventual link to paper*.

There are five main functions in this package: 

- `PICmodel.fit`: fits our Prevalence-Incidence-Cure model
- `PICmodel.predict`: makes predictions from the model
- `PICmodel.simulator`: simulates data under user-specified parameter values and covariates
- `score.test.gamma`: performs a score test whether the shape parameter of the Gamma distribution for progression is equal to one or not. 
- `simulator.gamma`: simulates data where progression follows a Gamma rather than exponential distribution and so can have a shape parameter not equal to one. Does not allow for covariates, yet. 


## Installation

You can install the most recent version of `PICmodel` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("kelsikroon/PICmodel")
```

## Examples

This is a basic example in a setting without covariates which illustrates how to simulate data, fit the model, make predictions from the plot and compare to a non-parametric cumulative incidence curve:

```{r example, messages=F, warning=F, out.width='50%',fig.align='center'}
library(PICmodel)
sim.thetas <- c(-5, -1.6, -1.2, 0.25)
sim.dat <- PICmodel.simulator(3000, c(), c(), c(), sim.thetas, show_prob = 0.9, interval=3, include.h=T)
head(sim.dat) # view simulated data

sim.fit <- PICmodel.fit(c(), c(), c(), sim.dat) # fit model to simulated data
sim.fit$summary # view model fit summary

sim.predict <- PICmodel.predict(data=sim.dat[1,], time.points = seq(0, 15, 0.5), fit=sim.fit)

library(survival) # compare model fit to non-parametric Kaplan-Meier curve 
sim.km.fit <- survfit(Surv(sim.dat$left, sim.dat$right, type='interval2')~1)

# plot PICmodel predictions and KM-curve to compare 
plot(sim.predict[[1]]$Time, sim.predict[[1]]$CR, type='l', xlab='time', ylab='CR', ylim=c(0.2, 0.65))
lines(sim.km.fit$time, 1-sim.km.fit$surv, col='blue')
```

To add covariates to the model we specify them separately for the progression, clearance, and prevalence parameters. For example if we wanted to add HPV16 as a covariate for progression and abnormal cytology as a covariate for prevalence then we would d the following: 
```{r example2, messages=F, warning=F, out.width='50%',fig.align='center'}

sim.thetas.cov <- c(-5, -1.6, 1, -1.2, -3, 4)
sim.dat2 <- PICmodel.simulator(5000, c("hpv"), c(), c("cyt"), sim.thetas.cov, show_prob = 0.9, interval=3, include.h=T)
head(sim.dat2) # view simulated data

sim.fit2 <- PICmodel.fit(c("hpv"), c(), c("cyt"), sim.dat2) # fit model to simulated data
sim.fit2$summary # view model fit summary

sim.predict2 <- PICmodel.predict(data=data.frame(hpv = c(1, 1, 0, 0), cyt=c(1, 0, 1, 0)), 
                              time.points = seq(0, 15, 0.5), fit=sim.fit2)

# compare model fit to non-parametric Kaplan-Meier curve 
sim.km.fit1 <- survfit(Surv(left, right, type='interval2')~1, data = sim.dat2[sim.dat2$hpv==1 & sim.dat2$cyt==1,])
sim.km.fit2 <- survfit(Surv(left, right, type='interval2')~1, data = sim.dat2[sim.dat2$hpv==1 & sim.dat2$cyt==0,])
sim.km.fit3 <- survfit(Surv(left, right, type='interval2')~1, data = sim.dat2[sim.dat2$hpv==0 & sim.dat2$cyt==1,])
sim.km.fit4 <- survfit(Surv(left, right, type='interval2')~1, data = sim.dat2[sim.dat2$hpv==0 & sim.dat2$cyt==0,])

```

```{r example3, messages=F, warning=F, out.width='50%', results='hold', fig.show='hold'}
# plot PICmodel predictions and KM-curve to compare 
plot(sim.predict2[[1]]$Time, sim.predict2[[1]]$CR, type='l', xlab='time', ylab='CR', ylim=c(0,1), col='darkblue')
lines(sim.km.fit1$time, 1-sim.km.fit1$surv, col='lightblue', lty=2)

lines(sim.predict2[[2]]$Time, sim.predict2[[2]]$CR, col = 'darkgreen')
lines(sim.km.fit2$time, 1-sim.km.fit2$surv, col='darkolivegreen1', lty=2)

lines(sim.predict2[[3]]$Time, sim.predict2[[3]]$CR,col='deeppink2')
lines(sim.km.fit3$time, 1-sim.km.fit3$surv, col='pink1', lty=2)

lines(sim.predict2[[4]]$Time, sim.predict2[[4]]$CR, col='darkorange3')
lines(sim.km.fit4$time, 1-sim.km.fit4$surv, col='orange1', lty=2)

legend("bottomright",
         legend=c("HPV16-pos & abnormal cyt", "HPV16-pos & normal cyt", "HPV16-neg & abnormal cyt", "HPV16-neg & normal cyt"),
         col=c("darkblue", "darkgreen", "deeppink2", "darkorange3"), lty=1, lwd=2, bty = "n", cex=0.75)
```


## Authors

- **Kelsi R. Kroon** k.kroon@amsterdamumc.nl
