---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  warning = FALSE, message = FALSE
)
```


# PICmodel

<!-- badges: start -->
<!-- badges: end -->

This `R` package fits prevalence-incidence-cure (PIC) models to interval-censored data to estimate the cumulative disease risk in a population with a temporarily elevated risk of disease, e.g., risk of CIN2+ in HPV-positive women. 

In longitudinal screening studies it is possible to observe prevalent, early or late events during follow-up. In our model, early events are modelled via a competing risks framework, infections either progress to the disease state or to a (latent) “cure” state (i.e. viral clearance) at constant rates. Late events are modelled by adding background risk to the model. Parameters can depend on individual risk factors and are estimated with an expectation-maximisation (EM) algorithm with weakly informative Cauchy priors. More details are given in the accompanying paper: *eventual link to paper*.

There are five main functions in this package: 
- `model.fit`
- `model.predict`
- `model.simulator`
- `score.test.gamma`
- `simulator.gamma`


## Installation

You can install the development version of PICmodel from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("kelsikroon/PICmodel")
```

## Examples

This is a basic example which shows you how to solve a common problem:

```{r example, messages=F, warning=F, out.width='50%',fig.align='center'}
library(PICmodel)
sim.thetas <- c(-5, -1.6, -1.2, 0.25)
sim.dat <- model.simulator(3000, c(), c(), c(), sim.thetas, show_prob = 0.9, interval=3, include.h=T)
head(sim.dat) # view simulated data
sim.fit <- model.fit(c(), c(), c(), sim.dat) # fit model to simulated data
sim.fit$summary # view model fit summary
sim.predict <- model.predict(c(), c(), c(), data=sim.dat[1,], time.points = seq(0, 15, 0.5), fit=sim.fit)


library(survival) # compare model fit to non-parametric Kaplan-Meier curve 
sim.km.fit <- survfit(Surv(sim.dat$left, sim.dat$right, type='interval2')~1)

# plot PICmodel predictions and KM-curve to compare 
plot(sim.predict[[1]]$Time, sim.predict[[1]]$CR, type='l', xlab='time', ylab='CR', ylim=c(0.2, 0.65))
lines(sim.km.fit$time, 1-sim.km.fit$surv, col='blue')
```
